Using high-throughput sequencing of ITS2 to describe *Symbiodinium* metacommunities in St. John, U.S. Virgin Islands

### 2017-06-05 supp/STJ2012_supp.Rmd
cleanup supp

---
title: "*Symbiodinium* communities in reef corals of St. John, U.S.V.I. (2012)"
subtitle: "Data analysis report"
author: "Ross Cunning"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 3, digits = 9)
```

@@

```{r STJmap, echo=F, message=F, cache=T, fig.height=3, fig.width=4, fig.align="center", fig.cap="Sampling locations in St. John"}
source("R/STJ_map.R")
```

@@

1. __100% OTU clustering.__ All ITS2 sequences from all samples were clustered at 100% identity using *uclust* and assigned taxonomy by BLAST to the [custom reference database](../data/ITS2db_trimmed.fasta).
2. __97% OTU clustering.__ All ITS2 sequences from all samples were clustered at 97% identity using *uclust* and the most abundant sequence within each OTU was selected as the representative sequence and assigned taxonomy by BLAST to the [custom reference database](../data/ITS2db_trimmed.fasta).
3. __97% OTU clustering within samples.__ Sequences from each individual sample were clustered separately at 97% similarity and the most abundant sequence from each cluster was selected as the representative sequence and assigned taxonomy by BLAST to the [custom reference database](../data/ITS2db_trimmed.fasta). OTUs with identical representative sequences were then merged across samples to create a global OTU table. By clustering at 97% _within_ samples, as opposed to across the entire dataset, OTUs from different samples that have _different representative sequences_ (i.e., different most abundant sequence within the 97% cluster) will remain separate even if these sequences are >97% similar to each other. This approach is used because _Symbiodinium_ ITS2 sequences are know to be multi-copy and intragenomically variable, and distinct _Symbiodinium_ types may contain some of the same ITS2 sequence variants in different relative proportions. Therefore, selecting the most abundant sequence variant _within_ a sample prevents a potentially distinct _Symbiodinium_ type from being collapsed into an OTU with a different representative sequence that happens to be more abundant in other samples within the dataset. For more discussion on this approach, see [this entry in my lab notebook](http://jrcunning.github.io/labnotebook//2016/02/12/ITS2_clustering.html).

@@

tax97 <- read.nw("data/otus_97/nw_tophits.tsv")
tax100 <- read.nw("data/otus_100/nw_tophits.tsv")
tax97bs <- read.nw("data/otus_97_bysample/nw_tophits.tsv")
tax97bsp <- read.nw("data/otus_97_byspecies/nw_tophits.tsv")

@@

tax97 <- collapse.idents(tax97)
tax97bs <- collapse.idents(tax97bs)
tax97bsp <- collapse.idents(tax97bsp)
tax100 <- collapse.idents(tax100)

@@

#tax97bsp <- get.st(tax97bsp)
tax97bsp <- with(tax97bsp, tax97bsp[order(otu, -sim), ])
tax97bsp <- tax97bsp[!duplicated(tax97bsp$otu), ]
rownames(tax97bsp) <- tax97bsp$otu

@@

```{r build_phyloseq}
# Import OTU tables
otu97 <- otu_table(read.table("data/otus_97/97_otus.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
otu97bs <- otu_table(read.table("data/otus_97_bysample/97_otus_bysample.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
otu97bsp <- otu_table(read.table("data/otus_97_byspecies/97_otus_byspecies.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
otu100 <- otu_table(read.table("data/otus_100/100_otus.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)

@@

# Build phyloseq objects
phy97 <- phyloseq(otu97, tax_table(as.matrix(tax97)), sam)
phy97bs <- phyloseq(otu97bs, tax_table(as.matrix(tax97bs)), sam)
phy97bsp <- phyloseq(otu97bsp, tax_table(as.matrix(tax97bsp)), sam)
phy100 <- phyloseq(otu100, tax_table(as.matrix(tax100)), sam)
```

@@

poormatch97bsp <- readLines("data/otus_97_byspecies/poormatch_IDs.txt")
poormatch97bsp <- data.frame(otu=str_extract(poormatch97bsp, "denovo[^ ]*"),
                             symbio=str_detect(poormatch97bsp, "Symbiodinium"),   # TRUE if Symbiodinium
                             stringsAsFactors = FALSE)
notsymbio97bsp <- poormatch97bsp[which(poormatch97bsp$symbio==F), 1]

@@

# Remove OTUs that are not Symbiodinium
phy97.f <- prune_taxa(!taxa_names(phy97) %in% notsymbio97, phy97)
phy97bs.f <- prune_taxa(!taxa_names(phy97bs) %in% notsymbio97bs, phy97bs)
phy97bsp.f <- prune_taxa(!taxa_names(phy97bsp) %in% notsymbio97bsp, phy97bsp)
phy100.f <- prune_taxa(!taxa_names(phy100) %in% notsymbio100, phy100)

# Filter OTUs by minimum count
# Set threshold count
n <- 10
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]
taxa97bsp <- taxa_sums(phy97bsp.f)[which(taxa_sums(phy97bsp.f) >= n)] 
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)
phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
phy97bsp.f <- prune_taxa(names(taxa97bsp), phy97bsp.f)
phy100.f <- prune_taxa(names(taxa100), phy100.f)

# Filter samples by minimum count
# Set threshold number of reads
sn <- 200
# Remove samples with fewer reads than threshold
phy97.f <- prune_samples(sample_sums(phy97.f)>=sn, phy97.f)
phy97bs.f <- prune_samples(sample_sums(phy97bs.f)>=sn, phy97bs.f)
phy97bsp.f <- prune_samples(sample_sums(phy97bsp.f)>=sn, phy97bsp.f)
phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)

# Filter OTUs by minimum count again in case any dropped below threshold after filtering samples
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]
taxa97bsp <- taxa_sums(phy97bsp.f)[which(taxa_sums(phy97bsp.f) >= n)] 
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)
phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
phy97bsp.f <- prune_taxa(names(taxa97bsp), phy97bsp.f)
phy100.f <- prune_taxa(names(taxa100), phy100.f)

@@

tax_table(phy97.f) <- as.matrix(get.st(data.frame(tax_table(phy97.f), stringsAsFactors=FALSE)))
tax_table(phy97bs.f) <- as.matrix(get.st(data.frame(tax_table(phy97bs.f), stringsAsFactors=FALSE)))
tax_table(phy97bsp.f) <- as.matrix(get.st(data.frame(tax_table(phy97bsp.f), stringsAsFactors=FALSE)))
tax_table(phy100.f) <- as.matrix(get.st(data.frame(tax_table(phy100.f), stringsAsFactors=FALSE)))

@@

#### Filtered dataset
```{r phy.f_histograms, cache=T, fig.height=4, fig.width=8}
# Compute summary statistics
stats97.f <- data.frame(`97% OTUs`=t(phystats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs.f)), check.names=F)
stats97bsp.f <- data.frame(`97% within-species OTUs`=t(phystats(phy97bsp.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)

# Create and plot histograms
taxhist97 <- hist(log10(taxa_sums(phy97.f)), plot=F)
taxhist97bs <- hist(log10(taxa_sums(phy97bs.f)), plot=F)
taxhist97bsp <- hist(log10(taxa_sums(phy97bsp.f)), plot=F)
taxhist100 <- hist(log10(taxa_sums(phy100.f)), plot=F)
samhist97 <- hist(log10(sample_sums(phy97.f)), plot=F)
samhist97bs <- hist(log10(sample_sums(phy97bs.f)), plot=F)
samhist97bsp <- hist(log10(sample_sums(phy97bsp.f)), plot=F)
samhist100 <- hist(log10(sample_sums(phy100.f)), plot=F)

par(mfrow=c(2, 4), mar=c(3,3,1,1))
plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist97bs, col="black", main="97% sample OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist97bsp, col="black", main="97% species OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist97bs, col="black", main="97% sample OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist97bsp, col="black", main="97% species OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)

# Create stats table
knitr::kable(cbind(stats97.f, stats97bs.f, stats97bsp.f, stats100.f))
```

#### Raw dataset
```{r phy_stats, cache=T, fig.height=4, fig.width=8}
# Compute summary statistics
stats97 <- data.frame(`97% OTUs`=t(phystats(phy97)), check.names=F)
stats97bs <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs)), check.names=F)
stats97bsp <- data.frame(`97% within-species OTUs`=t(phystats(phy97bsp)), check.names=F)
stats100 <- data.frame(`100% OTUs`=t(phystats(phy100)), check.names=F)

@@

taxhist97bsp <- hist(log10(taxa_sums(phy97bsp)), plot=F)
samhist97bsp <- hist(log10(sample_sums(phy97bsp)), plot=F)

@@

par(mfrow=c(2, 4), mar=c(3,3,1,1))
plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist97bs, col="black", main="97% within-sample OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist97bsp, col="black", main="97% within-species OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist97bs, col="black", main="97% within-sample OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist97bs, col="black", main="97% within-species OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)

# Create stats table
knitr::kable(cbind(stats97, stats97bs, stats97bsp, stats100))
```

### Transform count data
Count data are transformed to both relative abundance (proportions) and square-root proportions for downstream statistical analyses.
```{r transform}
# Convert to proportion (relative abundance)
phy97.f.p <- transform_sample_counts(phy97.f, function(x) x/sum(x))
phy97bs.f.p <- transform_sample_counts(phy97bs.f, function(x) x/sum(x))
phy97bsp.f.p <- transform_sample_counts(phy97bsp.f, function(x) x/sum(x))
phy100.f.p <- transform_sample_counts(phy100.f, function(x) x/sum(x))
# Apply transformation function
transform <- function(x) sqrt(x/sum(x))  # Set transformation function
phy97.f.t <- transform_sample_counts(phy97.f, transform)  # Transform data
phy97bs.f.t <- transform_sample_counts(phy97bs.f, transform) 
phy97bsp.f.t <- transform_sample_counts(phy97bsp.f, transform) 
phy100.f.t <- transform_sample_counts(phy100.f, transform)
```

# Clade overview
Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore. Each segment of the bar represents an individual OTU and is colored by clade. This plot is a nice overview of the whole dataset but only provides coarse information at the clade level.
```{r cladecomp, cache=T, fig.height=3, fig.width=4}
cladeAbund <- aggregate(data.frame(RelAbund=rowSums(otu_table(phy97bs.f.p))),
                        by=list(Clade=data.frame(tax_table(phy97bs.f.p))$Clade), FUN=sum)
cladeAbund$Prop <- prop.table(cladeAbund$RelAbund)

bars <- barplot(cladeAbund$Prop*100, col=taxcolors, space=0,
                names.arg=cladeAbund$Clade, xlab=expression(paste(italic('Symbiodinium'), " Clade")),
                ylab="Relative abundance (%)")
text(bars, cladeAbund$Prop*100+2, labels=round(cladeAbund$Prop*100, 1), xpd=T)

cladeAbund$Notus <- table(data.frame(tax_table(phy97bs.f.p))$Clade)

```{r cladecomp_bysample, cache=T, fig.height=6, fig.width=4}
par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
# Plot composition of 97% within-sample OTUs colored by OTU
#composition(phy97bs.f.p, col=rainbow(ntaxa(phy97bs.f.p)), legend=F)

@@

## *Diploria strigosa*  
Notice how 97% OTUs and 97% within-sample OTUs are **NOT THE SAME!** Specifically, notice how in the 97% OTU dataset, all samples except two are dominated by the same OTU, B1. In the 97% within-sample OTU dataset, those samples are dominated by three different OTUs -- B1, [B1]^6, and [B1]^3. Look at the 100% OTU dataset to see the composition of unique sequence variants in these samples. Indeed, these samples are dominated by **different sequence variants**. However, in the 97% OTU approach, they have all been collapsed into the same OTU, while in the 97% within-sample clustering approach, samples with different dominant sequence variants are maintained as separate OTUs, even though these different sequence variants are more than 97% similar to each other. I suggest that the  within-sample clustering approach is a better way of treating *Symbiodinium* ITS2 sequence data.
```{r dstr, cache=T, fig.height=16, fig.width=8}
# Create subsetted phyloseq objects for Diploria strigosa
dstr97.f.p <- subset_samples(phy97.f.p, Species=="strigosa")
dstr97bs.f.p <- subset_samples(phy97bs.f.p, Species=="strigosa")
dstr97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="strigosa")
dstr100.f.p <- subset_samples(phy100.f.p, Species=="strigosa")
# Plot custom barplots for Diploria strigosa
par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
otubarplot(dstr97.f.p, main="97% OTUs")
otubarplot(dstr97bsp.f.p, main="97% within-species OTUs")
otubarplot(dstr97bs.f.p, main="97% within-sample OTUs")
otubarplot(dstr100.f.p, main="100% OTUs")
```

@@

## *Porites furcata*
```{r pfur, cache=T, fig.height=16, fig.width=8}
# Create subsetted phyloseq objects for Porites furcata
pfur97.f.p <- subset_samples(phy97.f.p, Species=="furcata")
pfur97bsp.f.p <- subset_samples(phy97bsp.f.p, Species=="furcata")
pfur97bs.f.p <- subset_samples(phy97bs.f.p, Species=="furcata")
pfur100.f.p <- subset_samples(phy100.f.p, Species=="furcata")
# Plot custom barplots for Porites furcata
par(mfrow=c(4,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
otubarplot(pfur97.f.p, main="97% OTUs")
otubarplot(pfur97bsp.f.p, main="97% within-species OTUs")
otubarplot(pfur97bs.f.p, main="97% within-sample OTUs")
otubarplot(pfur100.f.p, main="100% OTUs")
```
```{r pfur.net, cache=T, fig.height=6, fig.width=6}
# Network analysis
pfur.net <- makenet(pfur97bs.f.p, 0)
par(mar=c(0,0,0,0))
set.seed(54538)
plotnet(pfur.net)
```

@@

# Differences between species
Whether *Symbiodinium* communities differ among species is evaluated using permutational analysis of variance (PERMANOVA).

>>>>

---
title: "Using high-throughput sequencing of ITS2 to describe *Symbiodinium* metacommunities in St. John, U.S. Virgin Islands"
subtitle: "Supplementary information and data analysis walk-through"
author: "Ross Cunning"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document: 
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
options(scipen = 3, digits = 9)
```

@@

```{r STJmap, echo=F, message=F, cache=T, fig.height=3, fig.width=4, fig.align="center", fig.cap="Sampling locations in St. John"}
library(rgdal)
stjmap <- readOGR("data/coastline/zipfolder", layer="Coastal_Coastline_line", verbose=F)  # units are in meters
stjmap <- spTransform(stjmap, CRS("+proj=longlat +datum=WGS84"))  # transform to lat/long

sitecoords <- data.frame(
  name=c("Booby Rock", "Tektite", "Kiddel", "Groot Pan", "Yawzi Point", "Anna's Point", "Haulover Bay"),
  lat=c(18.30327, 18.31057, 18.30770, 18.31067, 18.31515, 18.36710, 18.35008),
  lon=c(-64.71057, -64.72203, -64.71398, -64.71813, -64.72513, -64.73337, -64.67933)
)

viers <- c(-64.722697, 18.322277)

par(mar=c(0,0,0,0))
plot(stjmap, lwd=1, col="gray40",
     xlim=c(-64.75, -64.68), ylim=c(18.303, 18.370))
box()
text(-64.74, 18.34, expression(italic("       St. John\nU.S. Virgin Islands")), col="gray40")
axis(side=3, at=seq(-64.75, -64.65, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=3, at=seq(-64.75, -64.65, 0.04), line=-2, lwd=0, cex.axis=0.6)
axis(side=2, at=seq(18.3, 18.4, 0.01), line=0, tck=0.02, labels=FALSE)
axis(side=2, at=seq(18.3, 18.4, 0.02), line=-2, lwd=0, cex.axis=0.6)
with(sitecoords, {
  points(lon, lat, pch=21, bg="black")
  text(lon, lat, name, pos=c(4,2,4,4,2,1,3))
})
points(viers[1], viers[2], pch=2, col="gray40", cex=0.8)
text(viers[1], viers[2], expression(italic("VIERS")), col="gray40", pos=4, cex=0.7)
```

@@

1. __100% OTU clustering.__ All ITS2 sequences from all samples were clustered at 100% identity using *uclust* and assigned taxonomy by BLAST to the [custom reference database](../data/ITS2db_trimmed.fasta).
2. __97% OTU clustering.__ All ITS2 sequences from all samples were clustered at 97% identity using *uclust* and the most abundant sequence within each OTU was selected as the representative sequence and assigned taxonomy by BLAST to the [custom reference database](../data/ITS2db_trimmed.fasta).
3. __97% OTU clustering within samples.__ Sequences from each individual sample were clustered separately at 97% similarity and the most abundant sequence from each cluster was selected as the representative sequence and assigned taxonomy by BLAST to the [custom reference database](../data/ITS2db_trimmed.fasta). OTUs with identical representative sequences were then merged across samples to create a global OTU table. By clustering at 97% _within_ samples, as opposed to across the entire dataset, OTUs from different samples that have _different representative sequences_ (i.e., different most abundant sequence within the 97% cluster) will remain separate even if these sequences are >97% similar to each other. This approach is used because _Symbiodinium_ ITS2 sequences are know to be multi-copy and intragenomically variable, and distinct _Symbiodinium_ types may contain some of the same ITS2 sequence variants in different relative proportions. Therefore, selecting the most abundant sequence variant _within_ a sample prevents a potentially distinct _Symbiodinium_ type from being collapsed into an OTU with a different representative sequence that happens to be more abundant in other samples within the dataset.

@@

tax97 <- read.nw("data/otus_97/nw_tophits.tsv")
tax100 <- read.nw("data/otus_100/nw_tophits.tsv")
tax97bs <- read.nw("data/otus_97_bysample/nw_tophits.tsv")

@@

tax97 <- collapse.idents(tax97)
tax97bs <- collapse.idents(tax97bs)
tax100 <- collapse.idents(tax100)

@@

@@

```{r build_phyloseq}
# Import OTU tables
otu97 <- otu_table(read.table("data/otus_97/97_otus.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
otu97bs <- otu_table(read.table("data/otus_97_bysample/97_otus_bysample.tsv", header=T, check.names=F, row.names=1,
                              skip=1, sep="\t", comment.char=""), taxa_are_rows=T)
otu100 <- otu_table(read.table("data/otus_100/100_otus.tsv", header=T, check.names=F, row.names=1,
                               skip=1, sep="\t", comment.char=""), taxa_are_rows=T)

@@

# Build phyloseq objects
phy97 <- phyloseq(otu97, tax_table(as.matrix(tax97)), sam)
phy97bs <- phyloseq(otu97bs, tax_table(as.matrix(tax97bs)), sam)
phy100 <- phyloseq(otu100, tax_table(as.matrix(tax100)), sam)
```

@@

@@

# Remove OTUs that are not Symbiodinium
phy97.f <- prune_taxa(!taxa_names(phy97) %in% notsymbio97, phy97)
phy97bs.f <- prune_taxa(!taxa_names(phy97bs) %in% notsymbio97bs, phy97bs)
phy100.f <- prune_taxa(!taxa_names(phy100) %in% notsymbio100, phy100)

# Filter OTUs by minimum count
# Set threshold count
n <- 10
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)
phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
phy100.f <- prune_taxa(names(taxa100), phy100.f)

# Filter samples by minimum count
# Set threshold number of reads
sn <- 200
# Remove samples with fewer reads than threshold
phy97.f <- prune_samples(sample_sums(phy97.f)>=sn, phy97.f)
phy97bs.f <- prune_samples(sample_sums(phy97bs.f)>=sn, phy97bs.f)
phy100.f <- prune_samples(sample_sums(phy100.f)>=sn, phy100.f)

# Filter OTUs by minimum count again in case any dropped below threshold after filtering samples
# Identify OTUs below threshold count
taxa97 <- taxa_sums(phy97.f)[which(taxa_sums(phy97.f) >= n)]
taxa97bs <- taxa_sums(phy97bs.f)[which(taxa_sums(phy97bs.f) >= n)]
taxa100 <- taxa_sums(phy100.f)[which(taxa_sums(phy100.f) >= n)]
# Remove taxa below threshold count
phy97.f <- prune_taxa(names(taxa97), phy97.f)
phy97bs.f <- prune_taxa(names(taxa97bs), phy97bs.f)
phy100.f <- prune_taxa(names(taxa100), phy100.f)

@@

tax_table(phy97.f) <- as.matrix(get.st(data.frame(tax_table(phy97.f), stringsAsFactors=FALSE)))
tax_table(phy97bs.f) <- as.matrix(get.st(data.frame(tax_table(phy97bs.f), stringsAsFactors=FALSE)))
tax_table(phy100.f) <- as.matrix(get.st(data.frame(tax_table(phy100.f), stringsAsFactors=FALSE)))

@@

#### Filtered dataset
```{r phy.f_histograms, cache=T, fig.height=4, fig.width=8}
# Compute summary statistics
stats97.f <- data.frame(`97% OTUs`=t(phystats(phy97.f)), check.names=F)
stats97bs.f <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs.f)), check.names=F)
stats100.f <- data.frame(`100% OTUs`=t(phystats(phy100.f)), check.names=F)

# Create and plot histograms
taxhist97 <- hist(log10(taxa_sums(phy97.f)), plot=F)
taxhist97bs <- hist(log10(taxa_sums(phy97bs.f)), plot=F)
taxhist100 <- hist(log10(taxa_sums(phy100.f)), plot=F)
samhist97 <- hist(log10(sample_sums(phy97.f)), plot=F)
samhist97bs <- hist(log10(sample_sums(phy97bs.f)), plot=F)
samhist100 <- hist(log10(sample_sums(phy100.f)), plot=F)

par(mfrow=c(2, 3), mar=c(3,3,1,1))
plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist97bs, col="black", main="97% sample OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist97bs, col="black", main="97% sample OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)

# Create stats table
knitr::kable(cbind(stats97.f, stats97bs.f, stats100.f))
```

#### Raw dataset
```{r phy_stats, cache=T, fig.height=4, fig.width=8}
# Compute summary statistics
stats97 <- data.frame(`97% OTUs`=t(phystats(phy97)), check.names=F)
stats97bs <- data.frame(`97% within-sample OTUs`=t(phystats(phy97bs)), check.names=F)
stats100 <- data.frame(`100% OTUs`=t(phystats(phy100)), check.names=F)

@@

@@

par(mfrow=c(2, 3), mar=c(3,3,1,1))
plot(taxhist97, col="black", main="97% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(taxhist97bs, col="black", main="97% within-sample OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)

plot(taxhist100, col="black", main="100% OTU counts", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. OTUs", cex.lab=0.75, cex.axis=0.75)
plot(samhist97, col="black", main="97% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)
plot(samhist97bs, col="black", main="97% within-sample OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)

plot(samhist100, col="black", main="100% OTU reads per sample", xlim=c(0, 6), las=1, mgp=c(2,0.5,0),
     xlab="No. sequences (log10)", ylab="No. samples", cex.lab=0.75, cex.axis=0.75)

# Create stats table
knitr::kable(cbind(stats97, stats97bs, stats100))
```

### Transform count data
Count data are transformed to both relative abundance (proportions) and square-root proportions for downstream statistical analyses.
```{r transform}
# Convert to proportion (relative abundance)
phy97.f.p <- transform_sample_counts(phy97.f, function(x) x/sum(x))
phy97bs.f.p <- transform_sample_counts(phy97bs.f, function(x) x/sum(x))
phy100.f.p <- transform_sample_counts(phy100.f, function(x) x/sum(x))
# Apply transformation function
transform <- function(x) sqrt(x/sum(x))  # Set transformation function
phy97.f.t <- transform_sample_counts(phy97.f, transform)  # Transform data
phy97bs.f.t <- transform_sample_counts(phy97bs.f, transform) 
phy100.f.t <- transform_sample_counts(phy100.f, transform)
```

# Clade overview
Here the composition of each sample is plotted as a horizontal bar, sorted by species and shore. Each segment of the bar represents an individual OTU and is colored by clade. This plot is a nice overview of the whole dataset but only provides coarse information at the clade level.

```{r cladecomp_bysample, cache=T, fig.height=6, fig.width=4}
par(mfrow=c(1,1), mar=c(2, 1.5, 2, 5), lwd=0.1, cex=0.7)
# Plot composition of 97% within-sample OTUs colored by clade
composition(phy97bs.f.p, col=taxcolors[factor(data.frame(tax_table(phy97bs.f.p))[order(data.frame(tax_table(phy97bs.f.p))$Subtype), ]$Clade, levels=c("A","B","C","D","F","G"))], legend=T)
```

@@

## *Diploria strigosa*

```{r dstr, cache=T, fig.height=12, fig.width=8}
# Create subsetted phyloseq objects for Diploria strigosa
dstr97.f.p <- subset_samples(phy97.f.p, Species=="strigosa")
dstr97bs.f.p <- subset_samples(phy97bs.f.p, Species=="strigosa")
dstr100.f.p <- subset_samples(phy100.f.p, Species=="strigosa")
# Plot custom barplots for Diploria strigosa
par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
otubarplot(dstr97.f.p, main="97% OTUs")
otubarplot(dstr97bs.f.p, main="97% within-sample OTUs")
otubarplot(dstr100.f.p, main="100% OTUs")
```

@@

## *Porites furcata*
```{r pfur, cache=T, fig.height=12, fig.width=8}
# Create subsetted phyloseq objects for Porites furcata
pfur97.f.p <- subset_samples(phy97.f.p, Species=="furcata")
pfur97bs.f.p <- subset_samples(phy97bs.f.p, Species=="furcata")
pfur100.f.p <- subset_samples(phy100.f.p, Species=="furcata")
# Plot custom barplots for Porites furcata
par(mfrow=c(3,1), mar=c(4,4,2,0), mgp=c(0,0.5,0), lwd=0.1)
otubarplot(pfur97.f.p, main="97% OTUs")
otubarplot(pfur97bs.f.p, main="97% within-sample OTUs")
otubarplot(pfur100.f.p, main="100% OTUs")
```
```{r pfur.net, cache=T, fig.height=6, fig.width=6}
# Network analysis
pfur.net <- makenet(pfur97bs.f.p, 0)
par(mar=c(0,0,0,0))
set.seed(54538)
plotnet(pfur.net)
```

@@
